Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-04 mr-review.Rmd
added 1st dag

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set [@buniello2019nhgri].

@@

There are several ways that real data may violate the three MR assumptions. These include .

Recognition that the assumptions may be violated has inspired methods advances that 
enable assumption relaxations under certain conditions.

>>>>

Recent developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set, along with the sample size [@buniello2019nhgri].

@@

There are several ways that experimental data may violate one or more of the three MR assumptions.
Let's first consider the exclusion restriction assumption. It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

```{r, dag1, fig.cap = "Scenario where the genetic variant affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
g1 <- dagitty::dagitty( "dag {
	G -> W -> X -> Y
	W -> Y
}")
plot(dagitty::graphLayout(g1))
```

Recognition of the many ways that the MR assumptions may be violated has both inspired methods advances that 
enable assumption relaxations under certain conditions and motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
